Location and Scale-Invariant Power Transformations for Transforming Data to Normality

### 2025-10-29 manuscript_latex/manuscript.tex
Update appendix d.

\begin{table}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 10000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         7.65  &         6.96  &         7.28  &         6.58  &         7.45  &         6.78 \\
conventional                          & &         3.10  &         5.98  &         5.42  &         6.16  &         6.49  &         6.10 \\
conventional (z-score norm.)          & &         4.19  &         6.28  &         4.85  &         6.33  &         4.38  &         5.85 \\
conventional (robust scaling)         & &         4.20  &         6.28  &         4.78  &         6.33  &         4.30  &         5.86 \\
Raymaekers-Rousseeuw                  & &         4.46  &         3.55  &         5.39  &         3.69  &         6.29  &         5.84 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         6.21  &         3.83  &         4.95  &         3.85  &         4.51  &         3.50 \\
Raymaekers-Rousseeuw (robust scaling) & &         6.12  &         3.82  &         4.97  &         3.85  &         4.52  &         3.50 \\
invariant                             & & \textbf{2.99} &         5.55  & \textbf{3.65} &         6.01  & \textbf{3.52} &         5.57 \\
robust invariant                      & &         6.09  & \textbf{2.75} &         3.71  & \textbf{2.20} &         3.55  & \textbf{2.00} \\

>>>>

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Box-Cox transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. The clean dataset consists of sequences derived through inverse Box-Cox transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
If necessary, each sequence was shifted so that every element had a strictly positive value.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_box_cox}
\begin{tabular}{l | l r r r r r r}

@@

none                                  & &         7.63  &         6.95  &         7.29  &         6.59  &         7.47  &         6.78 \\
conventional                          & &         3.11  &         5.98  &         5.41  &         6.18  &         6.51  &         6.10 \\
conventional (z-score norm.)          & &         4.15  &         6.27  &         4.83  &         6.32  &         4.34  &         5.84 \\
conventional (robust scaling)         & &         4.20  &         6.29  &         4.84  &         6.31  &         4.35  &         5.84 \\
Raymaekers-Rousseeuw                  & &         4.50  &         3.56  &         5.39  &         3.69  &         6.30  &         5.80 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         6.23  &         3.80  &         4.93  &         3.85  &         4.48  &         3.50 \\
Raymaekers-Rousseeuw (robust scaling) & &         6.12  &         3.80  &         4.99  &         3.87  &         4.53  &         3.51 \\
invariant                             & & \textbf{2.96} &         5.59  & \textbf{3.62} &         6.01  & \textbf{3.51} &         5.58 \\
robust invariant                      & &         6.10  & \textbf{2.76} &         3.68  & \textbf{2.19} & \textbf{3.51} & \textbf{2.00} \\

\subsection{Ranking Yeo-Johnson transformations: residual
errors}\label{ranking-yeo-johnson-transformations-residual-errors}

In the main manuscript, transformation methods are compared by ranking
within each experiment, defined by the combination of sequence and
absence/presence of outliers. This allows for direct comparison of
methods between different experiments.

Table \ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
shows the average normalised residual error for these transformation
methods. This metric is computed from the residual error and the
residual error of the central portion (\(\kappa = 0.80\)) for sequences
without and with outliers, respectively. In each experiment, the
respective residual error was normalised by dividing with the sequence
length, and can be interpreted as the average residual error of a single
sample. This reduces the effect of sequence length on the residual
error.

Table
\ref{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
shows the average normalised difference in residual error between the
best method for each experiment and the individual methods. As above,
normalisation was done by division by the length of the sequence
corresponding to each experiment.

Both tables show that the differences between best-performing methods on
each dataset are relatively small.

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 100000 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.202 & 0.112 & 0.128 & 0.080 & 0.129 & 0.080 \\
conventional                          & & 0.058 & 0.094 & 0.085 & 0.078 & 0.129 & 0.080 \\
conventional (z-score norm.)          & & 0.061 & 0.094 & 0.091 & 0.073 & 0.091 & 0.073 \\
conventional (robust scaling)         & & 0.058 & 0.094 & 0.090 & 0.073 & 0.090 & 0.073 \\
Raymaekers-Rousseeuw                  & & 0.071 & 0.055 & 0.092 & 0.279 & 0.129 & 0.080 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.087 & 0.057 & 0.097 & 0.060 & 0.097 & 0.060 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.071 & 0.056 & 0.095 & 0.059 & 0.095 & 0.059 \\
invariant                             & & 0.057 & 0.099 & 0.083 & 0.075 & 0.083 & 0.075 \\
robust invariant                      & & 0.069 & 0.066 & 0.084 & 0.058 & 0.084 & 0.058 \\

\begin{table}
\begin{center}
\caption{
Comparison of average difference in residual error between Yeo-Johnson
transformation methods and the best transformation method within each
experiment, normalised by the corresponding sequence length. This is based on
either the residual error (without outliers) or the residual error of the
central portion (with outliers; $\kappa = 0.80$) over 3 datasets with 100000
sequences each. Residual errors were normalised by sequence length to reduce the
effect of sequence length on the residual error. The clean dataset consists of
sequences derived through inverse Yeo-Johnson transformation of data sampled
from a standard normal distribution. The dirty dataset contains sequences
sampled from asymmetric generalised normal distributions, centred at 0. The
shifted dataset also contains sequences sampled from asymmetric generalised
normal distributions, but centred at 100, and scaled by 0.001. Several
transformation methods include normalisation before transformation, indicated by
z-score normalisation (norm.) or robust scaling. A lower difference is better,
with a minimum value of 0.0. Discrepancies with the rank-based method may occur,
as normalised differences in residual error are not fully comparable between
experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_avg_residual_diff}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.150 & 0.068 & 0.050 & 0.033 & 0.049 & 0.033 \\
conventional                          & & 0.005 & 0.049 & 0.006 & 0.032 & 0.049 & 0.033 \\
conventional (z-score norm.)          & & 0.008 & 0.049 & 0.013 & 0.026 & 0.011 & 0.025 \\
conventional (robust scaling)         & & 0.005 & 0.050 & 0.012 & 0.026 & 0.010 & 0.026 \\
Raymaekers-Rousseeuw                  & & 0.018 & 0.011 & 0.014 & 0.233 & 0.049 & 0.033 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.034 & 0.012 & 0.018 & 0.014 & 0.017 & 0.013 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.018 & 0.011 & 0.017 & 0.012 & 0.015 & 0.011 \\
invariant                             & & 0.004 & 0.054 & 0.005 & 0.029 & 0.003 & 0.028 \\
robust invariant                      & & 0.016 & 0.021 & 0.006 & 0.011 & 0.004 & 0.010 \\

\subsection{Ranking Yeo-Johnson transformations: effect of sequence
length}\label{ranking-yeo-johnson-transformations-effect-of-sequence-length}

The main manuscript shows the overall ranking of different
Yeo-Johnson-based power transformation methods across three sets of
problems. To assess the effect of limited sequence lengths, we perform
ranking on subsets of the experiment. The results for ranking the subset
of sequences with 30 samples or fewer (\(n = 16001\)) is shown in Table
\ref{tab:comparison_methods_simulations_yeo_johnson_10_30}. In the clean
dataset without outliers, where all sequences were sampled from a normal
distribution and then underwent an inverse Yeo-Johnson transformation,
conventional and location- and scale-invariant are ranked similarly,
with no clear advantage to any method. For the remaining datasets, the
robust location- and scale-invariant method ranked best. It slightly
outperformed (robustly scaled) conventional Yeo-Johnson transformations
and (z-score normalised) robust transformations
\citep{Raymaekers2024-zf}, in datasets without and with outliers,
respectively.

In the subset for larger sequence lengths, with at least 1000 samples
per sequence (\(n = 33037\)), the cost of making the location- and
scale-invariant transformation robust against outliers becomes more
apparent (Table
\ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000}). In the
clean dataset without outliers, down-weighting of the tails of the
distribution by the robust location- and scale-invariant method leads to
larger residuals compared to other methods (Table
\ref{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}.

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Each sequence contained 30 samples or fewer.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30}
\begin{tabular}{l | l r r r r r r}

none                                  & &         7.75  &         6.81  &         7.27  &         6.40  &         7.45  &         6.60 \\
conventional                          & &         4.19  &         5.47  &         4.48  &         5.78  &         6.55  &         5.94 \\
conventional (z-score norm.)          & &         4.36  &         5.12  &         5.01  &         4.99  &         4.35  &         4.70 \\
conventional (robust scaling)         & & \textbf{4.05} &         5.63  &         4.38  &         5.49  &         3.76  &         5.20 \\
Raymaekers-Rousseeuw                  & &         5.36  &         4.19  &         5.14  &         4.62  &         6.32  &         5.83 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.56  &         4.08  &         5.67  &         4.05  &         5.03  &         3.81 \\
Raymaekers-Rousseeuw (robust scaling) & &         5.33  &         4.14  &         5.15  &         4.15  &         4.55  &         3.94 \\
invariant                             & &         4.11  &         5.92  &         4.12  &         5.67  &         3.62  &         5.37 \\
robust invariant                      & &         4.28  & \textbf{3.65} & \textbf{3.79} & \textbf{3.85} & \textbf{3.37} & \textbf{3.61} \\

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
Each sequence contained 30 samples or fewer. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_10_30_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.246 & 0.173 & 0.186 & 0.156 & 0.187 & 0.157 \\
conventional                          & & 0.139 & 0.142 & 0.142 & 0.132 & 0.187 & 0.157 \\
conventional (z-score norm.)          & & 0.139 & 0.142 & 0.142 & 0.130 & 0.142 & 0.130 \\
conventional (robust scaling)         & & 0.139 & 0.142 & 0.141 & 0.130 & 0.141 & 0.130 \\
Raymaekers-Rousseeuw                  & & 0.211 & 0.151 & 0.183 & 1.515 & 0.187 & 0.157 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.295 & 0.161 & 0.175 & 0.156 & 0.175 & 0.156 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.209 & 0.155 & 0.172 & 0.144 & 0.172 & 0.144 \\
invariant                             & & 0.141 & 0.144 & 0.140 & 0.132 & 0.140 & 0.132 \\
robust invariant                      & & 0.141 & 0.125 & 0.140 & 0.119 & 0.140 & 0.119 \\

\begin{table}
\begin{center}
\caption{
Comparison of average rank between Yeo-Johnson transformation methods based on either residual error (without outliers) or residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 33037 sequences each. Each sequence contained between 1000 and 10000 samples.
The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A rank of 1 is the best and a rank of 9 the worst. For each dataset, the best ranking transformation is marked in bold.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000}
\begin{tabular}{l | l r r r r r r}

none                                  & &         8.85  &         8.14  &         7.35  &         6.29  &         7.41  &         6.87 \\
conventional                          & &         3.50  &         5.65  &         3.52  &         7.78  &         6.42  &         6.05 \\
conventional (z-score norm.)          & &         5.24  &         5.94  &         7.34  &         4.98  &         6.26  &         4.79 \\
conventional (robust scaling)         & &         3.53  &         6.98  &         5.97  &         5.79  &         4.87  &         5.65 \\
Raymaekers-Rousseeuw                  & &         4.15  &         2.81  &         3.37  &         3.72  &         6.19  &         5.79 \\
Raymaekers-Rousseeuw (z-score norm.)  & &         5.34  & \textbf{1.86} &         6.57  &         3.39  &         5.49  &         3.30 \\
Raymaekers-Rousseeuw (robust scaling) & &         4.20  &         2.33  &         5.21  &         3.36  &         4.12  &         3.23 \\
invariant                             & & \textbf{2.36} &         7.59  & \textbf{2.66} &         6.82  & \textbf{1.98} &         6.81 \\
robust invariant                      & &         7.83  &         3.70  &         3.01  & \textbf{2.88} &         2.26  & \textbf{2.51} \\

\begin{table}
\begin{center}
\caption{
Comparison of average normalised residual error between Yeo-Johnson transformation methods based on either the residual error (without outliers) or the residual error of the central portion
(with outliers; $\kappa = 0.80$) over 3 datasets with 16001 sequences each. Residual errors were normalised by sequence length to reduce the effect of sequence length on the residual error.
Each sequence contained between 1000 and 10000 samples. The clean dataset consists of sequences derived through inverse Yeo-Johnson transformation
of data sampled from a standard normal distribution. The dirty dataset contains sequences sampled from asymmetric generalised normal distributions, centred at 0.
The shifted dataset also contains sequences sampled from asymmetric generalised normal distributions, but centred at 100, and scaled by 0.001.
Several transformation methods include normalisation before transformation, indicated by z-score normalisation (norm.) or robust scaling.
A lower normalised residual error is better. Discrepancies with the rank-based method may occur, as even with normalisation residual errors
are not fully comparable between experiments, whereas the ranks achieved within each experiment are.
}
\label{tab:comparison_methods_simulations_yeo_johnson_1000_10000_avg_residual}
\begin{tabular}{l | l r r r r r r}

none                                  & & 0.186 & 0.091 & 0.106 & 0.053 & 0.106 & 0.053 \\
conventional                          & & 0.014 & 0.076 & 0.060 & 0.057 & 0.106 & 0.053 \\
conventional (z-score norm.)          & & 0.021 & 0.076 & 0.070 & 0.050 & 0.070 & 0.050 \\
conventional (robust scaling)         & & 0.014 & 0.076 & 0.069 & 0.050 & 0.069 & 0.050 \\
Raymaekers-Rousseeuw                  & & 0.015 & 0.021 & 0.060 & 0.030 & 0.106 & 0.053 \\
Raymaekers-Rousseeuw (z-score norm.)  & & 0.021 & 0.020 & 0.070 & 0.030 & 0.070 & 0.030 \\
Raymaekers-Rousseeuw (robust scaling) & & 0.015 & 0.020 & 0.068 & 0.030 & 0.068 & 0.030 \\
invariant                             & & 0.013 & 0.084 & 0.059 & 0.055 & 0.059 & 0.055 \\
robust invariant                      & & 0.034 & 0.042 & 0.061 & 0.032 & 0.061 & 0.032 \\
